Palmer Archipelago (Antarctica) penguins: an alternative to Edgar Anderson's iris data

### 2020-06-27 manuscript/manuscript_text.Rmd
add pca & kmeans drafts

```{r setup, include=FALSE}
library(tidyverse)
library(here)
library(palmerpenguins)
library(kableExtra)
library(patchwork)
library(ggExtra)
library(gt)
library(ggpubr)
library(ggbeeswarm)
library(car)
knitr::opts_chunk$set(echo = TRUE)
```

@@

Data for each penguin species (Adélie, chinstrap and gentoo) were imported separately, then combined into a single table stored in the [*palmerpenguins*](https://allisonhorst.github.io/palmerpenguins/) package as object `penguins_raw`. Data in `penguins_raw` are in tidy format [@tidydata] and contain observations for 17 variables (Appendix Table 1) recorded completely or in part for 344 individual penguins.

@@

We feel that the `penguins` data is generally useful and approachable for statistics education, and that it is uniquely well-suited to replace `iris`. Comparisons presented here are representative examples, and are not exhaustive. Additional examples and tutorials are shared in the [user-contributed examples](https://allisonhorst.github.io/palmerpenguins/articles/articles/user_contributions.html) article on the package website.

@@

### Linear correlation & regression opportunities

#### Comparison 1: Linear trends across or within species

An overall approximately linear relationship can be explored using petal dimensions from `iris`, and penguin size (flipper length versus body mass) in `penguins` (Figure X). Educators may explore how linear regression across all species of iris or penguins (Figure X A,B) compares to regression performed separately for each species (Figure X C,D).

@@

iris_petal_scatter <- ggplot(iris, aes(x = Petal.Length, y = Petal.Width)) +
  geom_point(aes(color = Species, size = Species, shape = Species), size = 1.5, alpha = 0.7) +
  geom_smooth(aes(group = Species, color = Species), method = "lm", se = FALSE) +
  theme_minimal() +
  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
  theme(legend.position = c(0.3, 0.8)) +
  labs(x = "Petal length (cm)",
       y = "Petal width (cm)") +
  theme(plot.title.position = "plot")

iris_petal_scatter_all <- ggplot(iris, aes(x = Petal.Length, y = Petal.Width)) +
  geom_point(color = "gray50", size = 1.5, alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  theme_minimal() +
  labs(x = "Petal length (cm)",
       y = "Petal width (cm)",
       title = "Iris petal dimensions") +
  theme(plot.title.position = "plot")

@@

#### Comparison 2: linear relationships within species

More obvious clustering (by penguin species) exists when comparing bill length and depth (Figure X), and the relationship also provides an opportunity to introduce & explore Simpson's Paradox.

```{r, echo = FALSE, warning = FALSE, message = FALSE}
penguin_bill_scatter <- ggplot(penguins, aes(x = bill_length_mm, y = bill_depth_mm)) +
  geom_point(aes(color = species), size = 2, alpha = 0.7) +
  theme_minimal() +
  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
  theme(legend.position = "bottom") +
  labs(x = "Bill length (mm)", 
       y = "Bill depth (g)",
       title = "Penguin size", 
       color = "Species") +
  theme(plot.title.position = "plot")

iris_sepal_scatter <- ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(aes(color = Species), size = 2, alpha = 0.7) +
  theme_minimal() +
  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
  theme(legend.position = "bottom") +
  labs(x = "Sepal length (cm)",
       y = "Sepal width (cm)",
       title = "Iris sepal dimensions") +
  theme(plot.title.position = "plot")

penguin_bill_scatter + iris_sepal_scatter
```

@@

@@

### Clustering (e.g. k-means?)

### Classification (Alison Hill blog post?)

@@

A Web of Science search reveals that Fisher's 1936 paper has been cited in 7,258 publications (as of June 2020), most often in journals publishing in fields of Computer Science (57% of 7,258 citing references), Mathematics (49%), and Engineering (29%) - most of which we expect are crediting Anderson's *Iris* data, and not Fisher's eugenics work.

>>>>

```{r setup, include=FALSE}
library(tidyverse)
library(here)
library(palmerpenguins)
library(kableExtra)
library(patchwork)
library(ggExtra)
library(gt)
library(ggpubr)
library(ggbeeswarm)
library(car)
library(GGally)
library(recipes)
library(tidytext)
library(ggrepel)
library(shadowtext)
library(glue)
library(ggdendro)
knitr::opts_chunk$set(echo = TRUE)

theme_set(theme_minimal())
```

@@

For easier data access and use we created the [**palmerpenguins**](https://allisonhorst.github.io/palmerpenguins/) data package in *R*, which contains two data objects:

- `penguins_raw`: a table containing all raw data 
- `penguins`: a simplified table containing only size measurements, sex, island and species

The penguins data can also be imported directly from EDI, or accessed elsewhere (see *Penguin access for non-R users* section for more options).

### Uncurated `penguins_raw` data

Data for each penguin species (Adélie, chinstrap and gentoo) were imported from EDI, then combined into a single table stored in the [*palmerpenguins*](https://allisonhorst.github.io/palmerpenguins/) package as object `penguins_raw`. Data in `penguins_raw` are in tidy format [@tidydata] and contain observations for 17 variables (Appendix Table 1) recorded completely or in part for 344 individual penguins. For full variable descriptions and metadata, visit the data packages through EDI.

@@

We feel that the `penguins` data is generally useful and approachable for statistics education, and that it is uniquely well-suited to replace `iris`. Comparisons presented here are representative examples, and are not exhaustive. Additional examples and tutorials are shared in the [user-contributed examples](https://allisonhorst.github.io/palmerpenguins/articles/articles/user_contributions.html) article on the **palmerpenguins** package website.

@@

### Overview comparison: numeric variables

Relationships, distributions and correlations can be explored for the four numeric variables in `iris` (sepal width and length, petal width and length) and `penguins` (flipper length, body mass, bill length and depth) in pair plots (Figure X A,B).

```{r, warning = FALSE, message = FALSE, echo = FALSE, fig.align = "center"}
penguins %>%
  select(species, where(is.numeric)) %>% 
  ggpairs(aes(color = species),
          columns = c("flipper_length_mm", "body_mass_g", 
                      "bill_length_mm", "bill_depth_mm")) +
  scale_colour_manual(values = c("darkorange","purple","cyan4")) +
  scale_fill_manual(values = c("darkorange","purple","cyan4")) +
  theme_minimal() +
  theme(text = element_text(size = 10))

ggsave(here("figures","penguins_ggpairs.png"), height = 6, width = 5)

iris %>%
  ggpairs(aes(color = Species),
          columns = c("Petal.Length", "Petal.Width", 
                      "Sepal.Length", "Sepal.Width")) +
  scale_colour_manual(values = c("darkorange","purple","cyan4")) +
  scale_fill_manual(values = c("darkorange","purple","cyan4")) +
  theme_minimal() +
  theme(text = element_text(size = 10),
        panel.background = element_rect(fill = "gray90", color = "gray90"))

ggsave(here("figures","iris_ggpairs.png"), height = 6, width = 5)

### Linear correlation & regression

As seen in Figures (X,X) there are a number of opportunities to explore linear relationships and correlations in both `penguins` and `iris`. Here, we point out one comparison that is uniquely similar: 
petal dimensions (petal length versus petal width) from `iris`, and penguin size (flipper length versus body mass) in `penguins` (Figure X). We see that the overall trend across species is relatively linear for both `iris` and `penguins` (Figure X A,B). Teachers may encourage students explore how results of regression performed across all three species of iris or penguins (Figure X A,B) compare to results for regression performed within species (Figure X C,D).

Notably, distinctions between species are clearer for `iris` petals - particularly, the much smaller petals for *Iris setosa* - compared to `penguins`, in which Adélie and chinstrap penguins are largely overlapping in body size (body mass and flipper length), and are both generally smaller than gentoos. This difference is revisited in the k-means clustering example below.

```{r, echo = FALSE, warning = FALSE, message = FALSE, fig.align = "center"}

@@

iris_petal_scatter <- ggplot(iris, aes(x = Petal.Length, y = Petal.Width)) +
  geom_point(aes(color = Species, size = Species, shape = Species), size = 1.5, alpha = 0.7) +
  geom_smooth(aes(group = Species, color = Species), method = "lm", se = FALSE) +
  theme_minimal() +
  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
  theme(legend.position = c(0.3, 0.8)) +
  labs(x = "Petal length (cm)",
       y = "Petal width (cm)") +
  theme(plot.title.position = "plot",
        panel.background = element_rect(fill = "gray90", color = "gray90"))

iris_petal_scatter_all <- ggplot(iris, aes(x = Petal.Length, y = Petal.Width)) +
  geom_point(color = "gray50", size = 1.5, alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  theme_minimal() +
  labs(x = "Petal length (cm)",
       y = "Petal width (cm)",
       title = "Iris petal dimensions") +
  theme(plot.title.position = "plot",
        panel.background = element_rect(fill = "gray90", color = "gray90"))

@@

@@

### Linear discriminant analysis?

@@

The `iris` dataset appears frequently in examples for multivariate analyses, including principal component analysis (PCA). PCA is a dimensional reduction method that can be useful in capturing as much information about patterns in multivariate data as possible in a reduced number of dimensions (e.g., 2-dimensions for visualization). PCA with `iris` yields intuitive PC outcomes and clear clustering (Figure X) helpful for learners, as described in detail by others (ref, ref).

PCA with `penguins` reveals a similar outcome (Figure X): two somewhat overlapping species (Adélie and Chinstrap) with one clearly separated (Gentoo), and intuitive loadings that can be related to scatterplots in Figure X. Percent of total variance captured by the first and second principal components are very similar between `iris` and `penguins` (Figure X).

```{r, echo = FALSE, message = FALSE, warning = FALSE, fig.align = "center"}
# Copied from Alison Hill's post!

### PENGUINS PCA: 
penguin_recipe <-
  recipe(~., data = penguins) %>% 
  update_role(species, island, sex, new_role = "id") %>% 
  step_naomit(all_predictors()) %>% 
  step_normalize(all_predictors()) %>%
  step_pca(all_predictors(), id = "pca") %>% 
  prep()

penguin_pca <- 
  penguin_recipe %>% 
  tidy(id = "pca")

penguin_percvar <- penguin_recipe %>% 
  tidy(id = "pca", type = "variance") %>% 
  dplyr::filter(terms == "percent variance")

# penguin_pca

# Biplot:

# get pca loadings into wider format
pca_wider <- penguin_pca %>% 
  tidyr::pivot_wider(names_from = component, id_cols = terms)

# define arrow style:
arrow_style <- arrow(length = unit(.05, "inches"),
                     type = "closed")

# Make the biplot:
pca_plot <-
  juice(penguin_recipe) %>%
  ggplot(aes(PC1, PC2)) +
  geom_point(aes(color = species, shape = species), 
             alpha = 0.8, 
             size = 2) +
  scale_colour_manual(values = c("darkorange","purple","cyan4"))

penguins_biplot <- pca_plot +
  geom_segment(data = pca_wider,
               aes(xend = PC1, yend = PC2), 
               x = 0, 
               y = 0, 
               arrow = arrow_style) + 
  geom_shadowtext(data = pca_wider,
            aes(x = PC1, y = PC2, label = terms), 
            nudge_x = c(0.3,0.3,1.0,0.8), 
            nudge_y = c(-0.1,-0.2,0.1,-0.1),
            size = 3, 
            color = '#0A537D',
            bg.color = "white") +
  labs(title = "Penguins PCA biplot") +
  theme(legend.position = "bottom")

# For positioning (above): 
# 1: bill_length
# 2: bill_depth
# 3: flipper length
# 4: body mass

penguin_screeplot <- penguin_percvar %>% 
  ggplot(aes(x = component, y = value)) + 
  geom_col(fill = "#b6dfe2") + 
  xlim(c(0, 5)) + 
  ylim(c(0,80)) +
  ylab("% of total variance") +
  labs(title = "Penguins PCA screeplot") +
  geom_text(aes(label = round(value,2)), vjust=-0.25)

### IRIS PCA: 
iris_recipe <-
  recipe(~., data = iris) %>% 
  update_role(Species, new_role = "id") %>% 
  step_normalize(all_predictors()) %>%
  step_pca(all_predictors(), id = "pca") %>% 
  prep()

iris_pca <- 
  iris_recipe %>% 
  tidy(id = "pca")

# Biplot:

# get pca loadings into wider format
iris_wider <- iris_pca %>% 
  tidyr::pivot_wider(names_from = component, id_cols = terms)

# define arrow style:
arrow_style <- arrow(length = unit(.05, "inches"),
                     type = "closed")

# Make the biplot:
iris_pca_plot <-
  juice(iris_recipe) %>%
  ggplot(aes(PC1, PC2)) +
  geom_point(aes(color = Species, shape = Species), 
             alpha = 0.8, 
             size = 2) +
  scale_colour_manual(values = c("darkorange","purple","cyan4"))

iris_biplot <- iris_pca_plot +
  geom_segment(data = iris_wider,
               aes(xend = PC1, yend = PC2), 
               x = 0, 
               y = 0, 
               arrow = arrow_style) + 
  geom_shadowtext(data = iris_wider,
            aes(x = PC1, y = PC2, label = terms), 
            nudge_x = c(0.3,0.3,0.7,0.6), 
            nudge_y = c(-0.1,-0.2,0.1,-0.1),
            size = 3, 
            color = '#0A537D',
            bg.color = "white") +
  labs(title = "Iris PCA biplot") +
  theme(panel.background = element_rect(fill = "gray90", color = "gray90"),
        legend.position = "bottom")

iris_percvar <- iris_recipe %>% 
  tidy(id = "pca", type = "variance") %>% 
  dplyr::filter(terms == "percent variance")

iris_screeplot <- iris_percvar %>% 
  ggplot(aes(x = component, y = value)) + 
  geom_col(fill = "#b6dfe2") + 
  xlim(c(0, 5)) + 
  ylim(c(0,80)) +
  ylab("% of total variance") +
  labs(title = "Iris PCA screeplot") + 
  geom_text(aes(label = round(value,2)), vjust=-0.25) +
  theme(panel.background = element_rect(fill = "gray90", color = "gray90"))

(penguins_biplot | iris_biplot)/(penguin_screeplot | iris_screeplot)

ggsave(here("figures","pca_plots.png"), width = 8, height = 8)
```

### K-means clustering

Unsupervised clustering by k-means is a common and popular entryway to machine learning and classification, and frequently employs the `iris` data for introductory examples (refs). The `penguins` data provides similar opportunities for demonstrating, and illustrating limitations of, k-means clustering.

#### k-means clustering: two variables

First, we compare k-means clustering using two variables for each dataset: for `iris`, petal width and petal length, and for a subset only including male `penguins`, bill length and bill depth. For each, we designate three centroids (*k* = 3) since there are three species of both *Iris* and penguins.

Clustering by k-means with *Iris* petal dimension results in perfect separation for *Iris setosa* observations: all 50 *setosa* exist in a single cluster (cluster 1) with no *Iris virginica* or *versicolor* assigned to the cluster (Figure X, Table X). There is minimal species misclassification between the *versicolor* (cluster 2) and *versicolor* (cluster 3) species (Table X).

Similarly, Adélie penguins are highly separated from Chinstrap and Gentoo penguins by k-means clustering using only bill length and bill depth: of 73 Adélie penguins in the data, 70 are assigned to a single cluster (Cluster 2), and only 3 are assigned to the cluster containing most Gentoos (Cluster 3). No Chinstraps or Gentoos are assigned to Cluster 2 (Figure X). The distinction between Chinstraps and Gentoos is less clear, in agreement with overlap in bill dimensions between the species observed visually (Figure X).

```{r, echo = FALSE, message = FALSE, warning = FALSE, fig.align = "center"}
# TWO VARIABLE k-means comparison
# Male penguins: bill length vs. bill depth
# Iris: petal length vs. petal width

penguins_k <- penguins %>% 
  filter(sex == "male") %>% 
  select(species, bill_length_mm:body_mass_g) %>% 
  drop_na()

# Using bill_length & flipper_length
# THIS ONE SHOWS *pretty* good clustering

bills <- penguins_k %>% select(species, bill_length_mm, bill_depth_mm)

set.seed(1967)
kclust_bills <- kmeans(bills[,2:3], centers = 3, nstart = 20)
# kclust_bills

# Table of assigned clusters for penguins
bills_class <- table(kclust_bills$cluster,bills$species)%>% 
  as.data.frame() %>% 
  rename(cluster = Var1, species = Var2, n = Freq) %>% 
  pivot_wider(names_from = species, values_from = n)

# Same for iris:

petals <- iris %>% select(Species, Petal.Length, Petal.Width)

set.seed(1967)
kclust_petals <- kmeans(petals[,2:3], centers = 3, nstart = 25)
# kclust_petals

# Table of assigned clusters for iris petals: 
petal_class <- table(kclust_petals$cluster, petals$Species) %>% 
  as.data.frame() %>% 
  rename(cluster = Var1, Species = Var2, n = Freq) %>% 
  pivot_wider(names_from = Species, values_from = n)

bills_cluster <- cbind(bills, cluster_no = kclust_bills$cluster)

penguins_kmeans <- ggplot(data = bills_cluster, aes(x = bill_length_mm, 
                                     bill_depth_mm)) +
  geom_text(aes(label=cluster_no, color = species), 
                  size = 4) +
  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
  labs(title = "Penguins k-means clustering") +
  theme(legend.position = "bottom")

iris_cluster <-cbind(petals, cluster_no = kclust_petals$cluster)

iris_kmeans <- ggplot(data = iris_cluster, aes(x = Petal.Length, y = Petal.Width)) +
  geom_text(aes(label = cluster_no, color = Species), size = 4) +
  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) +
  theme(panel.background = element_rect(fill = "gray90", color = "gray90"),
        legend.position = "bottom") +
  labs(title = "Iris k-means clustering")

(penguins_kmeans | iris_kmeans)

# Full table of cluster assignments: 
combo_cluster <- cbind(bills_class, petal_class) %>% 
  kable(caption = "Table X. Assigned k-means clusters for results shown in Figure X") %>% 
  kable_styling(full_width = FALSE) %>% 
  add_header_above(c("Penguin bill cluster assignments" = 4, "Iris petal cluster assignments" = 4)) %>% 
  column_spec(c(5:8), background = "gainsboro")

combo_cluster
```
#### k-means clustering: all numeric variables

When all four numeric variables for each dataset are included in k-means clustering, *setosa* irises remain distinctly clustered apart from the other two iris species, but there is overlap across all penguin species (Table X).

```{r, echo = FALSE, warning = FALSE, message = FALSE, fig.align = "center"}
# FOUR VARIABLE k-means comparison

penguins_k_full <- penguins %>% 
  select(species, bill_length_mm:body_mass_g) %>% 
  drop_na()

set.seed(1967)
kclust_penguins <- kmeans(penguins_k_full[,2:5], centers = 3, nstart = 20)

# Table of assigned clusters for penguins
penguins_class <- table(kclust_penguins$cluster,penguins_k_full$species)%>% 
  as.data.frame() %>% 
  rename(cluster = Var1, species = Var2, n = Freq) %>% 
  pivot_wider(names_from = species, values_from = n)

# penguins_class
# Same for iris:

set.seed(1967)
iris_k_full <- kmeans(iris[,1:4], centers = 3, nstart = 25)

# Table of assigned clusters for iris petals: 
iris_class <- table(iris_k_full$cluster, iris$Species) %>% 
  as.data.frame() %>% 
  rename(cluster = Var1, Species = Var2, n = Freq) %>% 
  pivot_wider(names_from = Species, values_from = n)

# iris_class

# Full table of cluster assignments with all 4 numeric variables: 
combo_cluster_full <- cbind(penguins_class, iris_class) %>% 
  kable(caption = "Table X. Assigned k-means clusters for iris and penguins when including all four numeric variables for each") %>% 
  kable_styling(full_width = FALSE) %>% 
  add_header_above(c("Penguin cluster assignments" = 4, "Iris cluster assignments" = 4)) %>% 
  column_spec(c(5:8), background = "gainsboro")

combo_cluster_full
```

@@

Random, maybe include somewhere? A Web of Science search reveals that Fisher's 1936 paper has been cited in 7,258 publications (as of June 2020), most often in journals publishing in fields of Computer Science (57% of 7,258 citing references), Mathematics (49%), and Engineering (29%) - most of which we expect are crediting Anderson's *Iris* data, and not Fisher's eugenics work.

### Hierarchical clustering?

```{r}
hc_iris <- hclust(dist(iris[, 1:4]))
ggdendrogram(hc_iris, rotate = TRUE, size = 2)

p_cc <- penguins %>% 
  drop_na()

hc_penguins <- hclust(dist(p_cc[, 3:6]))
ggdendrogram(hc_penguins, rotate = TRUE, size = 2)
